Lifting CDCL to Relational Domains

### 2016-10-12 algorithm.tex
minor
conflict analysis

For a program with $N$ variables, let $L$ be the total number of 
meet irreducibles returned by a domain $D$.  For $D$ = {\em ItvDom}, the 
maximum value of $L$ is $2*N$, whereas the maximum value of $L$ is 
$2*(N^2)$ for $D$ = {\em OctDom}. Note that an octagon is the conjunction 
of all octagonal inequalities in the set $L$.
%\rmcmt{safety as satisfiability}
\subsection{Static Analyses Equations for Safety}
Currently, ACDCL handles loops in the program by unrolling the bounded loops. 
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKwFunction{decide}{Decide}
\SetKwFunction{deduce}{deduce}
\SetKwFunction{backtrack}{backtrack}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{nondet}{nondet}
\SetKwFunction{run}{run}
\SetKwFunction{learn}{learn}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{CONFLICT}
\SetKwData{safe}{safe}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of abstract deduction transformer, 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract domain $A$}
\Output{The status (\safe or \unsafe) and a counterexample if \unsafe}
$v \leftarrow \top$ \;
$\mathcal{T} \leftarrow \epsilon$ \;
$\mathcal{R} \leftarrow \emptyset$ \;
$result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
\uIf{$result$ == \conflict} {
  \return \safe \;
}
\While{$true$} 
{
  \uIf{$v$ is $\gamma$-complete} {
    \return \unsafe;
  }
  $d \leftarrow decide(v)$ \;
  $v \leftarrow v \meet d$ \; 
  $\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
  $\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
  $result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$\;
  \Do{$result == \conflict$} {
    learn() \;
    \uIf{$\neg$ conflict\_analysis()} {
      return \safe \;  
    }
    result $\leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $\langle ACDCL(S) \rangle$ \label{Alg:acdcl}}
\end{algorithm2e}
%
\Omit {
Static program analysis based on abstract
interpretation~\cite{DBLP:conf/emsoft/Cousot07} has been widely used to
verify certain classes of properties for safety-critical systems.  In
abstract interpretation, a given program is analysed with respect to a set
of given abstract domains.  However, ACDCL is a novel program analysis 
technique that embeds an abstract domain inside the CDCL algorithm.
From the abstract interpretation point of view, ACDCL is an abstract 
interpreter that uses decision and learning to increase transformer precision. 
From the decision procedure perspective, ACDCL is a SAT solver for program 
analysis constraints and is thus a strict generalisation of propositional CDCL solvers.
Constraint propagation in ACDCL uses fixed point iteration, decisions restrict the 
range of intervals and learning generates program analysis constraints (not assumptions) 
that preserve the error reachability.  Moreover, ACDCL implicitly 
perform program and property driven trace partitioning to increase the precision
of the analysis.  
}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The algorithm begins with a set $S$ of abstract deduction transformer 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract 
domain $A$, an abstract valuation $\langle v \leftarrow \top \rangle$, 
and an empty propagation trail and reason trail, $\mathcal{T}$, $\mathcal{R}$, 
respectively.  The propagation trail stores all meet irreducibles inferred 
by the abstact model search phase (deductions and decisions), while the 
reason trail maps elements of propagation trail to transformers that were 
used to drive them.

@@

>>>>

For a program with $N$ variables, let $L$ be the total number of 
meet irreducibles returned by a domain $D$.  For $D$ = {\em ItvDom}, the 
maximum value of $L$ is $2*N$, whereas the maximum value of $L$ is 
$2*(N^2)$ for $D$ = {\em OctDom}. Note that an octagon is the conjunction 
of all octagonal inequalities in the set $L$.
%\rmcmt{safety as satisfiability}
\subsection{Static Analysis Equations for Safety}
Static program analysis based on abstract interpretation~\cite{DBLP:conf/emsoft/Cousot07} 
perform safety analysis by computing fixed point to infer invariants 
over program variables.  However, bounded model checking (BMC) tries to search 
for a counterexample in a bounded execution trace, by symbolically executing a 
program up to a finite bound.  
Similar to BMC, ACDCL searches for a counterexample by solving the formula shown below.
For a set $N$ of program analysis constraints defined over a set 
of constraint variables $Var = \{X_i| i \in N\}$, representing a program
$\mathcal{P}$ and a concrete domain $\powerset(Env)$, the static analysis equation 
for safety of $\mathcal{P}$ is given as follows.  Here, $X_{init}, X_{Err}$ denote the initial valuation 
and the \rmcmt{final valuation} of constraint variables.
\[\varphi = X_{init} \subseteq Env \wedge \underset{p,q \in |\mathcal{P}|}
{\bigwedge} \{ X_p \subseteq post_{(p,q)}(X_q) \} \wedge X_{err} \supset \emptyset \] 
%
\begin{algorithm2e}[t]
\DontPrintSemicolon
\SetKwFunction{decide}{Decide}
\SetKwFunction{deduce}{deduce}
\SetKwFunction{backtrack}{backtrack}
\SetKwFunction{print}{print}
\SetKwFunction{return}{return}
\SetKwFunction{nondet}{nondet}
\SetKwFunction{run}{run}
\SetKwFunction{learn}{learn}
\SetKwFunction{continue}{continue}
\SetKwFunction{assign}{assign}
\SetKwRepeat{Do}{do}{while}
%\SetKwFunction{assume}{assume}
%\SetKwFunction{isf}{isFeasible}
\SetKwData{conflict}{CONFLICT}
\SetKwData{safe}{safe}
\SetKwData{unsafe}{unsafe}
\SetKwData{unknown}{unknown}
\SetKwData{true}{true}
\SetKwInOut{Input}{input}
\SetKwInOut{Output}{output}
\SetKwFor{Loop}{Loop}{}{}
\SetKw{KwNot}{not}
\begin{small}
\Input{Set $S$ of abstract deduction transformer, 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract domain $A$}
\Output{The status (\safe or \unsafe) and a counterexample if \unsafe}
$v \leftarrow \top$ \;
$\mathcal{T} \leftarrow \epsilon$ \;
$\mathcal{R} \leftarrow \emptyset$ \;
$result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
\uIf{$result$ == \conflict} {
  \return \safe \;
}
\While{$true$} 
{
  \uIf{$v$ is $\gamma$-complete} {
    \return \unsafe;
  }
  $d \leftarrow decide(v)$ \;
  $v \leftarrow v \meet d$ \; 
  $\mathcal{T} \leftarrow \mathcal{T} . d$ \; 
  $\mathcal{R}[|\mathcal{T}|] \leftarrow nil$ \;
  $result \leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$\;
  \Do{$result == \conflict$} {
    learn() \;
    \uIf{$\neg$ conflict\_analysis()} {
      return \safe \;  
    }
    result $\leftarrow \deduce(S,v,\mathcal{T},\mathcal{R})$ \;
  }
}
\end{small}
\caption{Abstract Conflict Driven Clause Learning $\langle ACDCL(S) \rangle$ \label{Alg:acdcl}}
\end{algorithm2e}
%
\subsection{Algorithm for Abstract Conflict Driven Clause Learning}
Algorithm~\ref{Alg:acdcl} presents an overview of the ACDCL algorithm.
The algorithm begins with a set $S$ of abstract deduction transformer 
$\widehat{ded_{\varphi}}$ for a formula $\varphi$ over an abstract 
domain $A$, an abstract valuation $\langle v \leftarrow \top \rangle$, 
and an empty propagation trail and reason trail, $\mathcal{T}$, $\mathcal{R}$, 
respectively.  The propagation trail stores all meet irreducibles inferred 
by the abstact model search phase (deductions and decisions), while the 
reason trail maps elements of propagation trail to transformers that were 
used to drive them.

@@

\paragraph \em{Solver State:}  The state of a ACDCL solver is a tuple 
of the form $\langle \mathcal{E}, S \rangle$.  Here, $\mathcal{E}$ 
is a sequence of labelled information of the form $(m,s)$ where 
$m$ is a meet irreducible and $s = decision$ if $m$ is a decision, 
or $s = deduction$ if $m$ is inferred by a deduction. And $S$ is 
a set of abstract deduction transformer.  A trail in SAT solver 
stores variable assignments of the form $(p, t)$, where $p$ is 
a propositional variable that appears at most once in the trail 
and $t$ is a truth assignment (true or false).  Whereas, a trail 
$\mathcal{T}$ in ACDCL contains a sequence of meet irreducibles 
inferred by deduction or decision phase where a variable in trail 
can be assigned multiple times, each time with increasingly precise 
bounds.  A current valuation $v$ is a meet of all elements of trail
$\mathcal{T}$, such that $v = \top$ when $\mathcal{T}$ is 
empty or $v=\underset{i \geq 0 \wedge i \leq |\mathcal{T}|}{\meet m_i}$, where 
$m_i \in \mathcal{T}$.  A solver is in {\em conflict} if some clauses in
$\mathcal{S}$ is not abstractly satisfied by $v$.

### 2016-10-12 decision.tex
minor changes
introduction
conflict analysis

\section{Decision}
A decision $\mathcal{D}$ is a meet irreducible that refines the 
current partial assignment.  A decision can be of the form 
$\langle x \preceq c \rangle$ or $\langle x \succeq c \rangle$, 
where $c$ is the bound and $x$ is a program variable.  A 
decision must always be consistent with respect to the 
trail $\mathcal{T}$, that is, it must not contradict with 
the elements in the trail.  Each variable in propositional CDCL 
is assigned at most once at a particular decision level.  However, 
a variable can be assigned several times in our lifting of CDCL to 
numerical domains, each time with increasingly precise bounds.

>>>>

\subsection{Decision}
A decision $\mathcal{D}$ is a meet irreducible that refines the 
current abstract valuation.  A decision can be of the form 
$\langle x \preceq c \rangle$ or $\langle x \succeq c \rangle$, 
where $c$ is the bound and $x$ is a program variable.  A 
decision must always be consistent with respect to the 
trail $\mathcal{T}$, that is, it must not contradict with 
the elements in the trail.

### 2016-10-12 domains.tex
conflict analysis

\section{Program and Abstract Domains}
In this section, we define a program and the notion of concrete domain 
and abstract domain. 
\subsection{Programs:}
%
A {\em program} $\mathcal{P}$ is formally defined as follows.
\[
\begin{array}[t]{@{}lll}
\mathcal{P} & {:}{:}{=} & \mathit{Procedure} \\
\mathit{Procedure} & {:}{:}{=} & \mathit{Statement} \mid \mathit{Procedure} \\
\mathit{Statement} & {:}{:}{=} & CStatement \mid \mathit{function(Var_1,\dots,Var_n)} \\
\mathit{CStatement} & {:}{:}{=} & x {:}{=} exp \mid \mathit{ITE}(b, s_1,s_2) \mid \mathit{s_1;s_2} \mid loop\{s\} \\
\end{array}
\]
Consider sets of expressions $Exp$ and Boolean expressions $BExp$
over variables $Var$ of $\mathcal{P}$.  The variables in $Var$ 
can take numeric values in $Val$.  A procedure is denoted as 
$\mathit{function(Var_1,\dots,Var_n)}$.  A $\mathit{CStatement}$ 
is an assignment, conditional, sequential concatenation or a loop. \\

\textbf{Control-flow Graph:} A CFG is a triple $(Loc, E, lbl)$, 
where $Loc$ is the set of locations with an unique start 
location $(init)$ and error location $(err)$, $E$ is the set 
of control-flow edges that are labelled with the set 
$lbl \in Statement$.  For the purpose of illustration, we 
assume that all the procedures are inlined.

\textbf{Program Analysis Constraints:}  A common approach to analyse 
a program $\mathcal{P}$ is to represent $\mathcal{P}$ as set of 
equations called, {\em program analysis constraints}.  The 
program analysis constraints, as shown below, is an assignment, 
or boolean expressions $BExp$ over program variables $Var$ or
a non-deterministic choice (ternary experssion) between statements $s_1$ and
$s_2$.  All the bounded loops are completely unrolled and represented using 
following constraints. Similarly, an assertion is of type $BExp$.  
\[c::= x:= exp | BExp | (exp) ? s_1, s_2\]
For example, figure~\ref{swssa} presents the program analysis constraints for 
a C program in Static Single Assignment (SSA) form.

>>>>

\section{Program and Abstract Domains}
In this section, we define a notion of program analysis constraints, concrete domain and abstract domain. 
\Omit {
\subsection{Programs:}
%
A {\em program} $\mathcal{P}$ is formally defined as follows.
\[
\begin{array}[t]{@{}lll}
\mathcal{P} & {:}{:}{=} & \mathit{Procedure} \\
\mathit{Procedure} & {:}{:}{=} & \mathit{Statement} \mid \mathit{Procedure} \\
\mathit{Statement} & {:}{:}{=} & CStatement \mid \mathit{function(Var_1,\dots,Var_n)} \\
\mathit{CStatement} & {:}{:}{=} & x {:}{=} exp \mid \mathit{ITE}(b, s_1,s_2) \mid \mathit{s_1;s_2} \mid loop\{s\} \\
\end{array}
\]
Consider sets of expressions $Exp$ and Boolean expressions $BExp$
over variables $Var$ of $\mathcal{P}$.  The variables in $Var$ 
can take numeric values in $Val$.  A procedure is denoted as 
$\mathit{function(Var_1,\dots,Var_n)}$.  A $\mathit{CStatement}$ 
is an assignment, conditional, sequential concatenation or a loop. \\

\textbf{Control-flow Graph:} A CFG is a triple $(Loc, E, lbl)$, 
where $Loc$ is the set of locations with an unique start 
location $(init)$ and error location $(err)$, $E$ is the set 
of control-flow edges that are labelled with the set 
$lbl \in Statement$.  For the purpose of illustration, we 
assume that all the procedures are inlined.  
}
\textbf{Program Analysis Constraints:}  A program $\mathcal{P}$ is represented as set of equations called, {\em program analysis constraints}.  The 
program analysis constraints, as shown below, is an assignment, 
or boolean expressions $BExp$ over program variables $Var$ or
a non-deterministic choice (ternary experssion) between statements $s_1$ and
$s_2$.  All the bounded loops are completely unrolled and represented using 
following constraints. Similarly, an assertion is of type $BExp$.  
\[c::= x:= exp | BExp | (exp) ? s_1, s_2\]
For example, figure~\ref{swssa} presents the program analysis constraints for 
a C program in Static Single Assignment (SSA) form.

### 2016-10-12 learning.tex
conflict analysis

\section{Clause Learning in Abstract Lattice}
Conflict graph for Intervals
Conflict graph for Octagons

@@

\textit{Backjumping:}
The backjumping level is defined by the literal of the conflict clause assigned
at the level that is the closest to the conflict one. In other words, the
backjumping level is the level closest to the root (decision level 0)  where the
conflict clause is still unit. If a conflict clause is "globally"  unit, then
the backjumping level is the root of the search tree.
1.DPLL style -- chronological backtracking \\
2. CDCL style -- non-chronological backtracking \\
  a. first-uip \\
  b. last-uip \\

\Omit{
**********************************************
\subsection{Lifting First UIP to Octagon domain}
**********************************************
unit-ness guarantee in octagon domain:
  Popped stmt: y23=1+y21
   Abstract value:
   D1: y23-y21 < 2 &&
   D2: y23+y21 > 0 &&
   D3: y21 < 1
   After backtracking, apply unit rule 
   y23 > 1 -- deduction from unit rule
   Value inconsistent !!

>>>>

\section{Abstract Conflict Analysis}
A conflict analysis in a SAT solver can be seen as abductive 
reasoning where a singleton assignment $c$ is replaced by a partial 
assignment that is sufficient to infer $c$.  Thus, a conflict 
analysis is a generalization of an underapproximation of a set of 
countermodels~\cite{sas,dhk2013-popl}, where a countermodel is a 
set of structures that do not satisfy a formula.  Since relational 
abstract domains have more complex structure than a partial 
assignments domain, so the lifting of {\em first-UIP} algorithm~\cite{uip} 
and clause learning to any arbitraty lattice structure is a non-trivial task.

The first-UIP algorithm in SAT solver works on a data structure called 
{\em implication graph} that computes a cut that suffices to produce a 
conflict.  The trail in ACDCL implicitly encodes a graph structure called 
{\em abstract conflict graph}, which records the dependencies between 
deductions made during abstract model search phase.  For example, consider a
formula $\varphi:= (x=5 \wedge y=x \wedge y=y+x \wedge y \leq 0)$ with current 
abstract valuation $v:= (x=[5,5])$.  
Figure~\ref{conflict} shows a snapshot of abstract implication graph that 
stores the deductions obtained from abstract model search using octagon domain.  
An arrow in the graph indicate that the predecessors of each node $n$ are 
sufficient to deduce $n$.  The last deduction conflict with the constraint 
$(y \leq 0)$.  Every cut in the graph, marked as {\em cut0, cut1}, provides a 
reason for conflict that can be used in learning.  An UIP is a special node such 
that any path from the last decision node to the conflict node must pass 
through it.  Unlike a SAT solver, a UIP in ACDCL may involve meet irreducibles 
that involves the same variables.  The first-UIP is a node closest to the 
conflict, and the last UIP is the decision node itself.  Computing UIPs ensures 
asserting cuts, that is, it yields clauses that generates new deduction after 
backtracking.  For our example, there exists multiple incomaparable reasons 
for conflict.  Here, cut0 is the first UIP.  Choosing cut0 yields a learnt 
clause $L_0: (x+y<15 \vee -x-y<-15 \vee -x+y<5 \vee x-y<-5 \vee y-10<0 \vee 
-y+10<0)$.  A learnt clause $\mathcal{L}$ is obtained by negating the reason 
for conflict.  
%
\begin{figure}
%\caption{An example of Octagon}\label{octagon}
\scalebox{.65}{\import{figures/}{conflict_graph.pspdftex}}
\label{conflict}
\end{figure} 
%    
\paragraph {\em Backjumping:}
A backjumping procedure undo all the assignments in the trail up to 
a decision level that restores the solver to a consistent state 
(non-conflicting).  The backjumping level is defined by the literal 
of the conflict clause that is closest to the root (decision level 0) 
where the conflict clause is still unit. If a conflict clause is 
globally unit, then the backjumping level is the root of the search tree.

The abstract clause learning and backjumping procedures in the abstract 
conflict graph is stated in terms of the state of ACDCL solver as follows. 
\[AbsLearn: \quad  (\mathcal{E},S) \rightarrow (\mathcal{E},S \wedge
\mathcal{L}) \quad \text{if} \; \mathcal{L} \notin S \; \textrm{and}
\; (S \wedge \mathcal{L}) \; \text{is not UNSAT} \]
\[AbsBackjump: \quad (\mathcal{E}_1(m,s)\mathcal{E}_2,S) \rightarrow
(\mathcal{E}_1,S) \quad \text{if} \; (\mathcal{E}_1,S) \; \text{is
consistent} \]

\subsection{Clause Learning in Abstract Lattice}
\Omit {
Conflict graph for Intervals
Conflict graph for Octagons

@@

1.DPLL style -- chronological backtracking \\
2. CDCL style -- non-chronological backtracking \\
  a. first-uip \\
  b. last-uip \\

**********************************************
\subsection{Lifting First UIP to Octagon domain}
**********************************************
unit-ness guarantee in octagon domain:
  Popped stmt: y23=1+y21
   Abstract value:
   D1: y23-y21 < 2 &&
   D2: y23+y21 > 0 &&
   D3: y21 < 1
   After backtracking, apply unit rule 
   y23 > 1 -- deduction from unit rule
   Value inconsistent !!
